---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BaMMI: Bayesian Microbiome-Metabolite Integration

BaMMI is an R package for Bayesian microbiome-metabolite data 
integration. It provides two models:

1. **BSRMM** - Bayesian sparse regression for microbiome-metabolite 
   data integration for univariate metabolite outcomes.
2. **MALMO** - Microbiome association with left-censored and missing 
   metabolic outputs for multivariate metabolite outcomes.

For a full tutorial with detailed examples and output, please see the
package vignette:
```{r, eval = FALSE}
browseVignettes("BaMMI")
```

## Installation

You can install the development version of BaMMI from GitHub with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Kai-Jiang-1/BaMMI")
```

## Example: BSRMM

```{r bsrmm_data, eval = FALSE}
library(BaMMI)

set.seed(123)
mydata <- gen_missing_value(
  n = 100, p = 200, snr = 1,
  proportion = 0.30, type = "mnar0.33"
)

bsrmm <- bsrmmgibbs(
  nburnin = 10000, niter = 20000,
  p = p, nop = nop,
  Y = Y, X = X, N = N,
  a = a, Q = mydata$Q, n = n
)
```

## Example: MALMO

> **Note:** Since MALMO is computationally expensive, we also provide
> MATLAB code at
> [https://github.com/Kai-Jiang-1/MALMO](https://github.com/Kai-Jiang-1/MALMO).
> We recommend running the MALMO model under a high performance
> computing environment for large datasets, for both the R package
> and MATLAB code.

```{r malmo_data, eval = FALSE}
set.seed(123)
mydata_mv <- gen_missing_value_mv(
  n = 50, p = 100, q = 3, snr = 1,
  proportion = 0.20, type = "mnar0.33"
)

result <- malmo_gibbs(
  Y = mydata_mv$Y_miss, X = mydata_mv$X,
  nburn = 10000, niter = 20000
)
```

## Citation

If you use BaMMI in your research, please cite the following papers:

**BSRMM** (accepted):
```
Jiang, K., et al. Bayesian sparse regression for microbiome-metabolite 
data integration with missing value imputation. 
Statistics in Medicine (Accepted, 2026).
```

**MALMO** (in preparation):
```
Jiang, K., et al. Microbiome association with left-censored
and missing metabolic outputs. (In Preparation, 2026).
```

## Contact

- **Author**: Kai Jiang
- **Email**: Kai.Jiang@uth.tmc.edu
- **GitHub**: https://github.com/Kai-Jiang-1/BaMMI
- **Institution**: UTHealth Houston
