---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BaMMI: Bayesian Microbiome-Metabolite Integration

BaMMI is an R package for Bayesian microbiome-metabolite data 
integration. It provides two models:

1. **BSRMM** - Bayesian sparse regression for microbiome-metabolite data integration for univariate metabolite outcomes. 
2. **MALMO** - Microbiome association with left-censored and missing metabolic outputs for multivariate metabolite outcomes.

## Installation

You can install the development version of BaMMI from GitHub with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Kai-Jiang-1/BaMMI")
```

## BSRMM

### Data Simulation

```{r bsrmm_data, eval = FALSE}
library(BaMMI)

set.seed(123)
mydata <- gen_missing_value(
  n          = 100,
  p          = 200,
  snr        = 1,
  proportion = 0.30,
  type       = "mnar0.33"
)

cat("True features:", which(mydata$coeffs[,1] != 0), "\n")
cat("Missing proportion:", mean(mydata$Y_miss == 0), "(1/3 MNAR + 2/3 MAR)\n")
```

### Run BSRMM Gibbs Sampler

```{r bsrmm_run, eval = FALSE}
## Prepare inputs
predictor <- mydata$X
outcome   <- mydata$Y_miss

## Standardize predictors
stand     <- list()
stand$mux <- colMeans(predictor)
stand$Sx  <- apply(predictor, 2, sd)
predictor <- apply(predictor, 2, function(x) (x - mean(x)) / sd(x))

## Train/test split
set.seed(33)
train_ind <- sort(sample(nrow(predictor),
                   size    = round(0.7 * nrow(predictor)),
                   replace = FALSE))

X <- as.matrix(predictor[train_ind, ])
Y <- as.matrix(outcome[train_ind, ])
n <- nrow(X); p <- ncol(X); c <- 100
N <- rbind(diag(p), matrix(c, 1, p))
Q <- mydata$Q; a <- rep(-12, p)
nop <- floor(n / 2)

## Run Gibbs sampler
bsrmm <- bsrmmgibbs(
  nburnin = 10000,
  niter   = 20000,
  p = p, nop = nop,
  Y = Y, X = X, N = N,
  a = a, Q = Q, n = n
)

## Posterior probabilities of inclusion
PPI      <- colMeans(bsrmm$gamma[(10000+2):(10000+20000+1), ])
selected <- which(PPI > 0.5)
cat("Selected features:", selected, "\n")

## Table of true vs selected features
table(
  True = ifelse(1:p %in% which(mydata$coeffs[,1] != 0), "Yes", "No"),
  Selected = ifelse(1:p %in% selected, "Yes", "No")
)
```

### Visualization

```{r bsrmm_plot, eval = FALSE, fig.width=10, fig.height=8}
## PPI plot
bsrmm_ppi(bsrmm, nburnin = 10000, niter = 20000)

## Forest plot
bsrmm_forest(bsrmm, nburnin = 10000, niter = 20000, stand = stand)

## MCMC diagnostics (multiple chains)
chain1 <- bsrmmgibbs(nburnin=10000, niter=20000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
chain2 <- bsrmmgibbs(nburnin=10000, niter=20000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
chain3 <- bsrmmgibbs(nburnin=10000, niter=20000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
chain4 <- bsrmmgibbs(nburnin=10000, niter=20000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
bsrmm_diagnostic(list(chain1, chain2, chain3, chain4), nburnin=10000, niter=20000)
```

## MALMO

> **Note:** Since MALMO is computationally expensive, we also provide
> MATLAB code at
> [https://github.com/Kai-Jiang-1/MALMO](https://github.com/Kai-Jiang-1/MALMO).
> We recommend running the MALMO model under a high performance
> computing environment for large datasets, for both the R package
> and MATLAB code. Here, we use a relatively small sample size to
> illustrate the model.

### Data Simulation

```{r malmo_data, eval = FALSE}
set.seed(123)
mydata_mv <- gen_missing_value_mv(
  n          = 50,
  p          = 20,
  q          = 2,
  snr        = 1,
  proportion = 0.30,
  type       = "mnar0.33"
)

cat("True features:", mydata_mv$p_true, "\n")
```

### Run MALMO Gibbs Sampler

```{r malmo_run, eval = FALSE}
result <- malmo_gibbs(
  Y       = mydata_mv$Y_miss,
  X       = mydata_mv$X,
  nburn   = 1000,
  niter   = 2000,
  display = FALSE
)

## Summarize C matrix
C_summary <- malmo_summarize(
  result$C_result,
  nburn = 1000,
  niter = 2000
)

## Selected features
selected <- which(rowSums(abs(C_summary$mean)) > 0)
cat("Selected features:", selected, "\n")
cat("True features:", mydata_mv$p_true, "\n")

## Table of true vs selected features
table(True = ifelse(1:nrow(C_summary$mean) %in% mydata_mv$p_true, "Yes", "No"),
      Selected = ifelse(1:nrow(C_summary$mean) %in% selected, "Yes", "No"))
```

### Visualization

```{r malmo_plot, eval = FALSE}
## Correlation heatmaps
malmo_plot_correlation(Y = mydata_mv$Y_true, X = mydata_mv$X)

## Venn/UpSet diagram
malmo_plot_venn(C_summary)

## MCMC diagnostics
malmo_diagnostic(result, nburn = 1000, niter = 2000)
```

## Contact

- **Author**: Kai Jiang
- **Email**: Kai.Jiang@uth.tmc.edu
- **GitHub**: https://github.com/Kai-Jiang-1/BaMMI
- **Institution**: UTHealth Houston
