---
title: "BaMMI Workflow: Bayesian Microbiome-Metabolite Integration"
author: "Kai Jiang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BaMMI Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 6,
  warning   = FALSE,
  message   = FALSE
)
```

## 1. Introduction

**BaMMI** (Bayesian Microbiome-Metabolite Integration) is an R package
for integrating microbiome and metabolite data using Bayesian methods.
It provides two main models:

- **BSRMM** (Bayesian Sparse Regression for Microbiome-Metabolite data):
  For univariate metabolite outcomes with missing value imputation.
  Performs simultaneous variable selection and missing value imputation
  using a Gibbs sampler.

- **MALMO** (Microbial Association with Left-censored Missing Metabolic
  Outputs): For multivariate metabolite outcomes using Bayesian reduced
  rank regression with missing value imputation.

### When to use BSRMM vs MALMO?

| | BSRMM | MALMO |
|--|-------|-------|
| **Outcome** | Univariate (single metabolite) | Multivariate (multiple metabolites) |
| **Coefficient** | Vector beta (p x 1) | Matrix C (p x q) |
| **Missing data** | MNAR/MAR/Mixed | MNAR/MAR/Mixed |
| **Rcpp** | No | Yes |
| **Speed** | Fast | Slow (HPC recommended) |

---

## 2. Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("Kai-Jiang-1/BaMMI")
```

```{r load}
library(BaMMI)
```

---

## 3. Project 1: BSRMM

### 3.1 Data Simulation

The `gen_missing_value()` function generates simulated microbiome
compositional data with correlated covariate structure and missing
values in the outcome. It supports four missing data mechanisms:
MNAR, MAR, mixed MNAR/MAR (1/3 + 2/3), and mixed MNAR/MAR (2/3 + 1/3).

```{r bsrmm_sim}
set.seed(123)
mydata <- gen_missing_value(
  n          = 100,   # number of observations
  p          = 200,   # number of microbiome features
  snr        = 1,     # signal to noise ratio
  proportion = 0.30,  # proportion of missing values
  type       = "mnar0.33"  # 1/3 MNAR + 2/3 MAR
)

# Check output structure
str(mydata)

# True signal features
cat("True features:", which(mydata$coeffs[, 1] != 0), "\n")
cat("Total true features:", sum(mydata$coeffs[, 1] != 0), "\n")
cat("Missing proportion:", mean(mydata$Y_miss == 0), "\n")
```

### 3.2 Data Preparation

Before running the Gibbs sampler, we need to:
1. Standardize the predictor matrix
2. Split into train/test sets
3. Set up model parameters

```{r bsrmm_prep}
## Prepare inputs
predictor <- mydata$X
outcome   <- mydata$Y_miss

## Standardize predictors
stand     <- list()
stand$mux <- colMeans(predictor)
stand$Sx  <- apply(predictor, 2, sd)
predictor <- apply(predictor, 2, function(x) (x - mean(x)) / sd(x))

## Train/test split (70/30)
set.seed(33)
train_ind <- sort(sample(nrow(predictor),
                   size    = round(0.7 * nrow(predictor)),
                   replace = FALSE))

predictor_train <- predictor[train_ind, ]
predictor_test  <- predictor[-train_ind, ]
outcome_train   <- outcome[train_ind, ]
outcome_test    <- outcome[-train_ind, ]

## Set up model parameters
X   <- as.matrix(predictor_train)
Y   <- as.matrix(outcome_train)
n   <- nrow(X)
p   <- ncol(X)
c   <- 100
N   <- rbind(diag(x = 1, nrow = p, ncol = p),
             matrix(data = c, nrow = 1, ncol = p))
Q   <- mydata$Q
a0  <- -12
a   <- rep(a0, p)
nop <- floor(n / 2)

cat("Training set: n =", n, ", p =", p, "\n")
cat("Test set: n =", nrow(predictor_test), "\n")
```

### 3.3 Running the BSRMM Gibbs Sampler

The `bsrmmgibbs()` function runs the Gibbs sampler for the BSRMM model.
For demonstration we use a small number of iterations. For real data
analysis, we recommend `nburnin = 10000` and `niter = 20000`.

```{r bsrmm_run}
set.seed(42)
bsrmm <- bsrmmgibbs(
  nburnin = 1000,
  niter   = 2000,
  p       = p,
  nop     = nop,
  Y       = Y,
  X       = X,
  N       = N,
  a       = a,
  Q       = Q,
  n       = n,
  display = FALSE
)

cat("Gibbs sampler completed!\n")
cat("Output components:", names(bsrmm), "\n")
```

### 3.4 Results

#### Posterior Probabilities of Inclusion (PPI)

```{r bsrmm_ppi}
## Compute PPI
nburnin <- 1000
niter   <- 2000
PPI      <- colMeans(bsrmm$gamma[(nburnin+2):(nburnin+niter+1), ])
selected <- which(PPI > 0.5)

cat("Selected features (PPI > 0.5):", selected, "\n")
cat("True features:", which(mydata$coeffs[, 1] != 0), "\n")

## Confusion table
tab <- table(
  Truth    = ifelse(1:p %in% which(mydata$coeffs[,1] != 0),
                    "True", "False"),
  Selected = ifelse(1:p %in% selected, "Yes", "No")
)
print(tab)

## Performance metrics
TP  <- sum(PPI > 0.5  & mydata$coeffs[, 1] == 1)
FP  <- sum(PPI > 0.5  & mydata$coeffs[, 1] == 0)
FN  <- sum(PPI <= 0.5 & mydata$coeffs[, 1] == 1)
TN  <- sum(PPI <= 0.5 & mydata$coeffs[, 1] == 0)
TPR <- TP / (TP + FN)
FPR <- FP / (FP + TN)
FDR <- ifelse((FP + TP) > 0, FP / (FP + TP), 0)
cat(sprintf("TP=%d | FP=%d | FN=%d | TN=%d\n", TP, FP, FN, TN))
cat(sprintf("TPR=%.3f | FPR=%.3f | FDR=%.3f\n", TPR, FPR, FDR))
```

#### PPI Plot

```{r bsrmm_ppi_plot, fig.width = 9, fig.height = 5}
bsrmm_ppi(bsrmm, nburnin = nburnin, niter = niter)
```

#### Forest Plot

```{r bsrmm_forest_plot, fig.width = 9, fig.height = 6}
bsrmm_forest(bsrmm, nburnin = nburnin, niter = niter, stand = stand)
```

### 3.5 MCMC Diagnostics

For MCMC diagnostics, we run multiple chains and compute the
Gelman-Rubin Rhat statistic. Values close to 1 indicate convergence.

```{r bsrmm_diag, fig.width = 10, fig.height = 8}
## Run 2 chains for demonstration
chain1 <- bsrmmgibbs(nburnin=1000, niter=2000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
chain2 <- bsrmmgibbs(nburnin=1000, niter=2000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)

diag_result <- bsrmm_diagnostic(
  bsrmm_list = list(chain1, chain2),
  nburnin    = 1000,
  niter      = 2000
)
```

---

## 4. Project 2: MALMO

> **Note:** Since MALMO is computationally expensive, we also provide
> MATLAB code at
> [https://github.com/Kai-Jiang-1/MALMO](https://github.com/Kai-Jiang-1/MALMO).
> We recommend running the MALMO model under a high performance
> computing environment for large datasets. Here, we use a small
> sample size to illustrate the model.

### 4.1 Data Simulation

The `gen_missing_value_mv()` function generates simulated microbiome
data with multivariate metabolite outcomes and missing values.

```{r malmo_sim}
set.seed(123)
mydata_mv <- gen_missing_value_mv(
  n          = 50,    # number of observations
  p          = 20,    # number of microbiome features
  q          = 3,     # number of metabolite outcomes
  snr        = 1,     # signal to noise ratio
  proportion = 0.20,  # proportion of missing values
  type       = "mnar0.33"
)

# Check output
str(mydata_mv)
cat("True features:", mydata_mv$p_true, "\n")
cat("Missing proportion:", mean(mydata_mv$Y_miss == 0), "\n")
cat("Y dimensions:", dim(mydata_mv$Y_miss), "\n")
cat("X dimensions:", dim(mydata_mv$X), "\n")
```

### 4.2 Running the MALMO Gibbs Sampler

```{r malmo_run}
result <- malmo_gibbs(
  Y       = mydata_mv$Y_miss,
  X       = mydata_mv$X,
  nburn   = 500,
  niter   = 1000,
  display = FALSE
)

cat("MALMO completed!\n")
cat("Output components:", names(result), "\n")
```

### 4.3 Results

#### Summarize Coefficient Matrix C

```{r malmo_summarize}
C_summary <- malmo_summarize(
  result$C_result,
  nburn = 500,
  niter = 1000
)

## Selected features
selected_mv <- which(rowSums(abs(C_summary$mean)) > 0)
cat("Selected features:", selected_mv, "\n")
cat("True features    :", mydata_mv$p_true, "\n")

## Confusion table
tab_mv <- table(
  Truth    = ifelse(1:nrow(C_summary$mean) %in% mydata_mv$p_true,
                    "True", "False"),
  Selected = ifelse(1:nrow(C_summary$mean) %in% selected_mv,
                    "Yes", "No")
)
print(tab_mv)
```

#### Correlation Heatmaps

```{r malmo_correlation, fig.width = 10, fig.height = 5}
malmo_plot_correlation(
  Y = mydata_mv$Y_true,
  X = mydata_mv$X
)
```

#### Heatmap of Selected Features

```{r malmo_heatmap, fig.width = 8, fig.height = 6}
malmo_plot_heatmap(C_summary)
```

#### Venn Diagram of Selected Features

```{r malmo_venn, fig.width = 7, fig.height = 7}
venn_result <- malmo_plot_venn(C_summary)
```

### 4.4 MCMC Diagnostics

```{r malmo_diag, fig.width = 10, fig.height = 8}
diag_mv <- malmo_diagnostic(
  result = result,
  nburn  = 500,
  niter  = 1000
)
```

---

## 5. Session Info

```{r session_info}
sessionInfo()
```
