---
title: "BaMMI Workflow: Bayesian Microbiome-Metabolite Integration"
author: "Kai Jiang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BaMMI Workflow: Bayesian Microbiome-Metabolite Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 6,
  warning   = FALSE,
  message   = FALSE
)
```

## 1. Introduction

**BaMMI** (Bayesian Microbiome-Metabolite Integration) is an R package for integrating microbiome and metabolite data using Bayesian methods. It provides two main models:

- **BSRMM** (Bayesian Sparse Regression for Microbiome-Metabolite Data Integration):
  For univariate metabolite outcomes with missing value imputation.
  Performs simultaneous variable selection and missing value imputation
  using a Gibbs sampler.

- **MALMO** (Microbial Association with Left-censored Missing Metabolic
  Outputs): For multivariate metabolite outcomes using Bayesian reduced
  rank regression with missing value imputation.

---

## 2. Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("Kai-Jiang-1/BaMMI")
```

```{r load}
library(BaMMI)
```

---

## 3. Project 1: BSRMM

### 3.1 Data Simulation

The `gen_missing_value()` function returns log-transformed microbial compositional data with a correlated covariance structure and a user-specified missing proportion for the outcomes. It supports four different missingness mechanisms, including pure MNAR, pure MAR, 1/3 MNAR + 2/3 MAR, and 2/3 MNAR + 1/3 MAR. Here, we demonstrate the 1/3 MNAR + 2/3 MAR scenario.

```{r bsrmm_sim}
set.seed(123)
mydata <- gen_missing_value(
  n          = 100,   # number of observations
  p          = 200,   # number of microbiome features
  snr        = 1,     # signal to noise ratio
  proportion = 0.30,  # proportion of missing values
  type       = "mnar0.33"  # 1/3 MNAR + 2/3 MAR
)

# True signal features
cat("True features:", which(mydata$coeffs[, 1] != 0), "\n")
cat("Total true features:", sum(mydata$coeffs[, 1] != 0), "\n")
cat("Missing proportion:", mean(mydata$Y_miss == 0), "\n")
```

### 3.2 Data Preparation

Before running the BSRMM model, we standardize the predictors, split the data into training and testing sets, and set up the model parameters. The training set is used to fit the BSRMM model, while the test set is used for evaluating predictive performance.

```{r bsrmm_prep}
## Prepare inputs
predictor <- mydata$X
outcome   <- mydata$Y_miss

## Standardize predictors
stand     <- list()
stand$mux <- colMeans(predictor)
stand$Sx  <- apply(predictor, 2, sd)
predictor <- apply(predictor, 2, function(x) (x - mean(x)) / sd(x))

## Train/test split (70/30)
set.seed(33)
train_ind <- sort(sample(nrow(predictor),
                   size    = round(0.7 * nrow(predictor)),
                   replace = FALSE))

predictor_train <- predictor[train_ind, ]
predictor_test  <- predictor[-train_ind, ]
outcome_train   <- outcome[train_ind, ]
outcome_test    <- outcome[-train_ind, ]

## Set up model parameters
X   <- as.matrix(predictor_train)
Y   <- as.matrix(outcome_train)
n   <- nrow(X)
p   <- ncol(X)
c   <- 100
N   <- rbind(diag(x = 1, nrow = p, ncol = p),
             matrix(data = c, nrow = 1, ncol = p))
Q   <- mydata$Q
a0  <- -12
a   <- rep(a0, p)
nop <- floor(n / 2)

cat("Training set: n =", n, ", p =", p, "\n")
cat("Test set: n =", nrow(predictor_test), "\n")
```

### 3.3 Running the BSRMM Gibbs Sampler

The `bsrmmgibbs()` function runs the Gibbs sampler for the BSRMM model.

```{r bsrmm_run}
set.seed(123)
bsrmm <- bsrmmgibbs(
  nburnin = 5000,
  niter   = 10000,
  p       = p,
  nop     = nop,
  Y       = Y,
  X       = X,
  N       = N,
  a       = a,
  Q       = Q,
  n       = n,
  display = FALSE
)
```

### 3.4 Results

#### Posterior Probabilities of Inclusion (PPI)

```{r bsrmm_ppi}
## Compute PPI
nburnin <- 5000
niter   <- 10000
PPI      <- colMeans(bsrmm$gamma[(nburnin+2):(nburnin+niter+1), ])
selected <- which(PPI > 0.5)

## Confusion table
cat("Confusion Table: Selected vs. True Features\n")
tab <- table(
  Truth    = ifelse(1:p %in% which(mydata$coeffs[,1] != 0),
                    "Yes", "No"),
  Selected = ifelse(1:p %in% selected, "Yes", "No")
)
print(tab)
```

### We provide two functions to visualize the results of BSRMM: `bsrmm_ppi()` for plotting the PPI of each feature, and `bsrmm_forest()` for plotting the posterior mean and credible intervals of the regression coefficients for the selected features.

#### PPI Plot

```{r bsrmm_ppi_plot, fig.width = 9, fig.height = 5}
bsrmm_ppi(bsrmm, nburnin = nburnin, niter = niter)
```

#### Forest Plot

```{r bsrmm_forest_plot, fig.width = 9, fig.height = 6}
bsrmm_forest(bsrmm, nburnin = nburnin, niter = niter, stand = stand)
```

### 3.5 MCMC Diagnostics

For MCMC diagnostics, we run multiple chains and compute the
Gelman-Rubin Rhat statistic. Values close to 1 indicate convergence.

```{r bsrmm_diag, fig.width = 10, fig.height = 8}
## Run 2 chains for demonstration
chain1 <- bsrmmgibbs(nburnin=5000, niter=10000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)
chain2 <- bsrmmgibbs(nburnin=5000, niter=10000, p=p, nop=nop,
                     Y=Y, X=X, N=N, a=a, Q=Q, n=n)

diag_result <- bsrmm_diagnostic(
  bsrmm_list = list(chain1, chain2),
  nburnin    = 5000,
  niter      = 10000
)
```

All parameters show Rhat values close to 1, indicating good convergence of the MCMC chains.
---

## 4. Project 2: MALMO

> **Note:** Since MALMO is computationally expensive, we also provide
> MATLAB code at
> [https://github.com/Kai-Jiang-1/MALMO](https://github.com/Kai-Jiang-1/MALMO).
> We recommend running the MALMO model under a high performance
> computing environment for large datasets. Here, we use a small
> sample size to illustrate the model.

### 4.1 Data Simulation

The `gen_missing_value_mv()` function generates multivariate metabolite outcomes with a user-specified missing proportion. Missing values are recorded as zeros. The predictors are generated from a log-transformed multivariate normal distribution with a correlated covariance structure. The function supports four different missingness mechanisms, including pure MNAR, pure MAR, 1/3 MNAR + 2/3 MAR, and 2/3 MNAR + 1/3 MAR. Here, we demonstrate the 1/3 MNAR + 2/3 MAR scenario.

```{r malmo_sim}
set.seed(123)
mydata_mv <- gen_missing_value_mv(
  n          = 50,    # number of observations
  p          = 20,    # number of microbiome features
  q          = 3,     # number of metabolite outcomes
  snr        = 1,     # signal to noise ratio
  proportion = 0.20,  # proportion of missing values
  type       = "mnar0.33"
)

cat("True features:", mydata_mv$p_true, "\n")
cat("Missing proportion:", mean(mydata_mv$Y_miss == 0), "\n")
cat("Y dimensions:", dim(mydata_mv$Y_miss), "\n")
cat("X dimensions:", dim(mydata_mv$X), "\n")
```

### 4.2 Running the MALMO Gibbs Sampler

The `malmo_gibbs()` function runs the Gibbs sampler for the MALMO model. 

```{r malmo_run}
result <- malmo_gibbs(
  Y       = mydata_mv$Y_miss,
  X       = mydata_mv$X,
  nburn   = 1000,
  niter   = 2000,
  display = FALSE
)
```

### 4.3 Results

#### Summarize Coefficient Matrix C

```{r malmo_summarize}
C_summary <- malmo_summarize(
  result$C_result,
  nburn = 1000,
  niter = 2000
)

## Confusion table
selected_mv <- which(rowSums(abs(C_summary$mean)) > 0)
tab_mv <- table(
  Truth    = ifelse(1:nrow(C_summary$mean) %in% mydata_mv$p_true,
                    "True", "False"),
  Selected = ifelse(1:nrow(C_summary$mean) %in% selected_mv,
                    "Yes", "No")
)
print(tab_mv)
```

### In this package, we include several functions to help users visualize results from MALMO, including exploratory analysis. The function `malmo_plot_correlation()` provides an overall overview of the correlations among outcomes and between outcomes and predictors. The function `malmo_plot_venn()` is designed to show the overlap of selected features across different outcomes. If the number of outcomes is greater than 3, the function returns an UpSet plot instead of a Venn diagram.

#### Correlation Heatmaps

```{r malmo_correlation, fig.width = 10, fig.height = 5}
malmo_plot_correlation(
  Y = mydata_mv$Y_true,
  X = mydata_mv$X
)
```


#### Venn Diagram of Selected Features

```{r malmo_venn, fig.width = 7, fig.height = 7}
venn_result <- malmo_plot_venn(C_summary)
```

### 4.4 MCMC Diagnostics

```{r malmo_diag, fig.width = 10, fig.height = 8}
diag_mv <- malmo_diagnostic(
  result = result,
  nburn  = 1000,
  niter  = 2000
)
```
An absolute Geweke z-score less than 1.96 indicates good convergence for the parameter.

---

## 5. Session Info

```{r session_info}
sessionInfo()
```
